---
title: "Data Visualization"
author: "Max Murphy"
format: revealjs
---
```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.align="center")

theme_set(theme_classic(base_size = 28))
```

# The Power of Data Visualization

---

## Introduction

- Data visualization is a vital tool for understanding and communicating insights from data
- In this presentation, we will explore the importance and utility of data visualization techniques


## Why Data Visualization?

- Humans are visual creatures; we process visual information more effectively than numbers or text
- Data visualization helps us comprehend complex patterns, relationships, and trends in data
- It enables us to make informed decisions and communicate findings with clarity


## Key Benefits of Data Visualization

1. **Insight Generation**: Visual representations can reveal hidden patterns and insights that may go unnoticed in raw data.
2. **Storytelling**: Visualization facilitates storytelling by presenting data in a compelling and memorable way.
3. **Simplification**: Complex data can be simplified and transformed into intuitive visual forms, aiding comprehension.


## Exploring Data with Visualizations

Let's explore the **gapminder dataset** using various visualization techniques.

- The gapminder dataset contains socio-economic indicators for different countries over time.
- It includes variables such as life expectancy, GDP per capita, population, and more.
- The dataset covers multiple years and several countries around the world.

---

## `gapminder` Dataset {auto-animate="true"}

```{r}
#| echo: true
#| eval: true

# Load the Iris dataset
data(gapminder, package = "gapminder")

# Display the first few rows of the dataset
head(gapminder)
```

## `gapminder` Dataset {auto-animate="true"}

- The dataset contains 1,704 observations and 6 variables
- The variables are:
    - **country**: Country name
    - **continent**: Continent name
    - **year**: Year
    - **lifeExp**: Life expectancy at birth
    - **pop**: Population
    - **gdpPercap**: GDP per capita

## Summary Statistics {auto-animate="true"}

```{r}
#| echo: true
#| eval: true

# Display summary statistics for the dataset
summary(gapminder)
```



## Exploring the Data {auto-animate="true"}
Let's explore the data and try to answer some questions:

## Exploring the Data {auto-animate="true"}
Let's explore the data and try to answer some questions:

- How does life expectancy relate to GDP per capita?

## Exploring the Data {auto-animate="true"}

- How does life expectancy relate to GDP per capita?

```{r}
#| echo: true
#| eval: false

# Plot life expectancy vs. GDP per capita
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    labs(x = "GDP per capita", y = "Life expectancy")
```

## Exploring the Data {auto-animate="true"}

- How does life expectancy relate to GDP per capita?

```{r}
#| echo: false
#| eval: true

# Plot life expectancy vs. GDP per capita
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    labs(x = "GDP per capita", y = "Life expectancy")
```

## Exploring the Data {auto-animate="true"}

There's a lot of data in this plot. Let's simplify it by looking at the data for a single year.

```{r}
#| echo: false
#| eval: true

# Plot life expectancy vs. GDP per capita
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    labs(x = "GDP per capita", y = "Life expectancy")
```

## Exploring the Data {auto-animate="true"}

There's a lot of data in this plot. Let's simplify it by looking at the data for a single year.

```{r}
#| echo: true
#| eval: false
gapminder_2007 <- gapminder |> filter(year == 2007)
```

## Exploring the Data {auto-animate="true"}
There's a lot of data in this plot. Let's simplify it by looking at the data for a single year.

```{r}
#| echo: true
#| eval: false
gapminder_2007 <- gapminder |> filter(year == 2007)

# Plot life expectancy vs. GDP per capita
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    labs(x = "GDP per capita", y = "Life expectancy")
```

## Exploring the Data {auto-animate="true"}
There's a lot of data in this plot. Let's simplify it by looking at the data for a single year.

```{r}
#| echo: false
#| eval: true
gapminder_2007 <- gapminder |> filter(year == 2007)

# Plot life expectancy vs. GDP per capita
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    labs(x = "GDP per capita", y = "Life expectancy")
```


## Exploring the Data {auto-animate="true"}

- How do population, GDP per capita, and life expectancy vary across countries?

## Exploring the Data {auto-animate="true"}

- How do population, GDP per capita, and life expectancy vary across countries?

```{.r code-line-numbers="|5"}
#| echo: true
#| eval: false

ggplot(gapminder_2007, 
    aes(x = gdpPercap, y = lifeExp, size = pop, color = continent)) +
    geom_point() +
    labs(x = "GDP per Capita", y = "Life Expectancy", 
         size = "Population", color = "Continent")
```

## Exploring the Data {auto-animate="true"}

- How do population, GDP per capita, and life expectancy vary across countries?

```{r}
#| echo: false
#| eval: true

ggplot(gapminder_2007, 
    aes(x = gdpPercap, y = lifeExp, size = pop, color = continent)) +
    geom_point() +
    labs(x = "GDP per Capita", y = "Life Expectancy", size = "Population", color = "Continent")
```

## Exploring the Data {auto-animate="true"}
We can also look at data over time by plotting year on the x-axis.

## Exploring the Data {auto-animate="true"}

- How does life expectancy vary over time for each country? for each continent?

## Exploring the Data {auto-animate="true"}

- How does life expectancy vary over time for each country? for each continent?

```{.r code-line-numbers="|4"}
#| echo: true
#| eval: false
ggplot(gapminder, 
  aes(x = year, y = lifeExp, color = continent, group = country)) +
  geom_line() +
  labs(x = "Year", y = "Life Expectancy", color = "Continent")
```

## Exploring the Data {auto-animate="true"}

```{r}
#| echo: false
#| eval: true 
ggplot(gapminder, aes(x = year, y = lifeExp, color = continent, group = country)) +
  geom_line() +
  labs(x = "Year", y = "Life Expectancy", color = "Continent")
```

## Conclusion

- Data visualization is an important tool for exploring data, communicating information, and answering questions
- It can reveal patterns, tell compelling stories, and allows us to make informed decisions