---
title: "Berkeley admission data"
# author: "Yutong Wang"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr) 
# devtools::install_github("rstudio-education/dsbox")
library(dsbox)
library(tidyverse)
library(scales) # plot labeling

tutorial_options(
  exercise.timelimit = 60,
  # A simple checker function that just returns the message in the check chunk
  exercise.checker = function(check_code, ...) {
    list(
      message = eval(parse(text = check_code)),
      correct = logical(0),
      type = "info",
      location = "append"
    )
  }
)
knitr::opts_chunk$set(error = TRUE)
```

## Welcome

In this tutorial, we will explore a Berkeley admission dataset and learn how to 

* visualize a dataset including categorical variables
* play around with a dataset, e.g. data wrangling, tidying, acquisition
* think critically as a statistician

The tutorial is based on [_datasciencebox.org_](https://datasciencebox.org/course-materials/_slides/u2-d16-simpsons-paradox/u2-d16-simpsons-paradox.html#3).

## Berkeley admission data

* Study carried out by the Graduate Division of the University of California, 
Berkeley in the early 70’s to evaluate whether there was a gender bias in 
graduate admissions.
* The data come from six departments. For confidentiality we'll call them A-F.
* We have information on whether the applicant was male or female and whether 
they were admitted or rejected.

```{r data1, exercise = TRUE, exercise.eval = FALSE}
head(ucbadmit, n = 5)
```

`head()` lets you take a quick look at the first part of a data-storing object. 
We can specify how many rows we'd like to see through the argument `n`. 

```{r data2, exercise = TRUE, exercise.eval = FALSE}
dim(ucbadmit)
```

`dim` retrieves the dimension of an object. How many applicants are there in the 
dataset? What information (features) do we know about each individual?

### Basic statistics 
```{r basic-stat1, exercise = TRUE, exercise.eval = FALSE}
table(ucbadmit$admit) 
```

```{r basic-stat3, exercise = TRUE, exercise.eval = FALSE}
table(ucbadmit$dept) 
```

```{r basic-stat4, exercise = TRUE, exercise.eval = FALSE}
table(ucbadmit$gender) 
```

## Question of interest

* In plain English: Is there any gender bias in the admission process, e.g. 
males are more likely to be admitted than females?
* Formulating with a bit statistics: Is the percentage of males admitted indeed 
higher than females?
* That is, to compare (1) the percentage of admitted males out of all the male 
applicants and (2) the percentage of admitted females out of all the female 
applicants.

```{r question1, exercise = TRUE, exercise.eval = FALSE}
tbl_perc <- table(ucbadmit$admit, ucbadmit$gender)
tbl_perc
```

solution 1:
```{r question2, exercise = TRUE, exercise.eval = FALSE}
tbl_perc <- table(ucbadmit$admit, ucbadmit$gender)
tbl_perc[2, 1] / (tbl_perc[1, 1] + tbl_perc[2, 1]) # female
tbl_perc[2, 2] / (tbl_perc[1, 2] + tbl_perc[2, 2]) # male
```

solution 2:
```{r question3, exercise = TRUE, exercise.eval = FALSE}
sum(ucbadmit$admit == "Admitted" & ucbadmit$gender == "Female") / sum(ucbadmit$gender == "Female")
sum(ucbadmit$admit == "Admitted" & ucbadmit$gender == "Male") / sum(ucbadmit$gender == "Male")
```

### Let's visualize it!

```{r gender-admit, fig.show = TRUE}
ggplot(ucbadmit, aes(y = gender, fill = admit)) + # specify the variables for the y axis and color
  geom_bar(position = "fill") + # we want a barplot
  labs(title = "Admit by gender", # add a plot title
       y = NULL, x = NULL) + # remove axis titles
  theme_classic() + theme(legend.position = "bottom") # nice aesthetics
```

## What if we look at it by department? 
```{r department, include = FALSE}
ucbadmit %>%
  count(dept, gender, admit) %>%
  pivot_wider(names_from = dept, values_from = n)
```

```{r department-plot, fig.show = TRUE}
ggplot(ucbadmit, aes(y = gender, fill = admit)) +
  geom_bar(position = "fill") +
  facet_wrap(. ~ dept) + # separate plots by department
  scale_x_continuous(labels = label_percent()) + # specify x axis label
  labs(title = "Admissions by gender and department",
       x = NULL, y = NULL, fill = NULL) +
  theme_classic() + theme(legend.position = "bottom") 
```

### Comparison

What do you notice when comparing the department-specific barplots with the 
barplot of the six departments combined?
```{r gender-admit-rep, fig.show = TRUE}
ggplot(ucbadmit, aes(y = gender, fill = admit)) + 
  geom_bar(position = "fill") + 
  labs(title = "Admit by gender", 
       y = NULL, x = NULL) +
  theme_classic() + theme(legend.position = "bottom")
```

## Simpson's Paradox

### A motivating example
```{r echo=FALSE, message=FALSE}
df <- tribble(
  ~x, ~y, ~z,
  2,   4,  "A",
  3,   3,  "A",
  4,   2,  "A",
  5,   1,  "A",
  6,   11, "B",
  7,   10, "B",
  8,   9,  "B",
  9,   8,  "B"
)
```

```{r echo = FALSE, fig.show = TRUE}
ggplot(df) +
  geom_point(aes(x = x, y = y), color = "darkgray", size = 5) +
  theme_minimal()
```

What if we try to fit the relationship between two variables on x and y axis?
```{r echo = FALSE, message = FALSE, fig.show = TRUE}
ggplot(data = df) +
  geom_point(aes(x = x, y = y), color = "darkgray", size = 5) +
  geom_smooth(aes(x = x, y = x), color = "darkgray") +
  theme_minimal()
```

What is there is actually a third variable we were not aware of before?!
```{r echo = FALSE, message = FALSE, fig.show = TRUE}
ggplot(data = df) +
  geom_point(aes(x = x, y = y, color = z), size = 5) +
  geom_smooth(aes(x = x, y = x), method = "lm", color = "darkgray") +
  theme_minimal()
```

```{r echo = FALSE, message = FALSE, fig.show = TRUE}
ggplot(data = df) +
  geom_point(aes(x = x, y = y, color = z), size = 5) +
  geom_smooth(aes(x = x, y = x), method = "lm", color = "darkgray") +
  geom_smooth(aes(x = x, y = y, color = z), method = "lm") +
  theme_minimal()
```

## Takeaways

- Not considering an important variable when studying a relationship can result in **Simpson's paradox**
- Simpson's paradox illustrates the effect that omission of an explanatory variable can have on the measure of association between another explanatory variable and a response variable
- The inclusion of a third variable in the analysis can change the apparent relationship between the other two variables
- [Video](https://www.youtube.com/watch?v=E_ME4P9fQbo)

### Questions
* What is the third variable in our analysis of the Berkeley admission data?
* Given the data (evidence) that we have, is there any gender bias in the 
admission process, e.g. males are more likely to be admitted than females?
* Is there any similar scenario or potential issue you could come up with in our 
real life?